### 2017-12-11 submission/manuscript.Rmd
incorporate edits to first results section -- adding clarity

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
Using the combined data set we first assessed whether there were any broad scale community differences that could be detected as disease severity worsened. Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). However, in tissue, this effect was not observed when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). We next tested whether these detectable differences in the community resulted in a significant increase in RR. Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`) and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)) [Figure 2]. Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)). Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

>>>>

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened. For feces and not tissue a decrease in evenness and Shannon diversity was associated with carcinoma [Figure 1]. Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). We next tested whether these detectable differences in community resulted in significant increases in RR. For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))` and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

### 2017-12-11 submission/manuscript.tex
incorporate edits to first results section -- adding clarity

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} Using the combined data set we first assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. Using power transformed and Z-score
normalized \(\alpha\)-diversity metrics, both evenness and Shannon
diversity in feces, but not tissue, were lower in those with carcinoma
{[}Figure 1{]}. Using linear mixed-effect models to control for study
and variable region, there was a significant decrease from control to
adenoma to carcinoma for both evenness (P-value = 0.025) and Shannon
diversity (P-value = 0.043). However, in tissue, this effect was not
observed when resampling of the same individual was also controlled for
(P-value \textgreater{} 0.05). We next tested whether these detectable
differences in the community resulted in a significant increase in RR.
Within fecal samples, a decrease in Shannon diversity and evenness
resulted in a significantly increased RR for carcinoma (P-value = 0.0047
and 0.00034, respectively) {[}Figure 2{]}. Although these values were
significant, the effect size was relatively small for both metrics
(Shannon RR = 1.33 (1.09 - 1.62) and evenness RR = 1.36 (1.15 - 1.61))
{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
even smaller than what was observed for carcinoma (RR = 1.16 (1.01 -
1.34)). Interestingly, for both adenoma and carcinoma there was no
increase in RR within tissue samples for any alpha diversity metric
investigated {[}Table S1-S3{]}.

>>>>

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} By power transforming and Z-score normalizing the
\(\alpha\)-diversity metrics for the entire data set we assessed whether
there were any broad scale community differences that could be detected
as disease severity worsened. For feces and not tissue a decrease in
evenness and Shannon diversity was associated with carcinoma {[}Figure
1{]}. Using linear mixed-effect models to control for study, resampling
of the same individual (tissue studies only), and variable region, there
was a significant decrease from control to adenoma to carcinoma for both
evenness and Shannon diversity in stool (P-value = 0.025 and 0.043,
respectively) but not for tissue (P-value \textgreater{} 0.05). We next
tested whether these detectable differences in community resulted in
significant increases in RR. For fecal samples, a decrease in evenness
resulted in a significantly increased RR for carcinoma (RR = 1.36 (1.15
- 1.61), P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value
= 0.032) while a decrease in Shannon diversity only increased the RR for
carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
Interestingly, for both adenoma and carcinoma there was no increase in
RR within tissue samples for any alpha diversity metric investigated
{[}Table S1-S3{]}.
